Local PCA results

### 2016-05-13 medicago/medicago_data_setup.Rmd
added in total variance output to eigen_windows and split out cov_pca

```

@@

To get windows out of the VCF file and recode them, we'll just pipe a call to `bcftools`.
Here's a dummy version that will pull out windows of 100 SNPs for only the first four individuals.
```{r win_fn}
get.indivs <- c("HM001","HM002","HM003","HM004")
bcf.file <- "data/chr1-filtered-set-2014Apr15.bcf"
# first read in the info about sites
bcf.con <- pipe(paste("bcftools query -f '%POS\\n' -r chr1:1-100000",bcf.file),open="r")
bcf.sites <- list( chr1=scan(bcf.con) )
close(bcf.con)
# now use bcftools to pull out sites we want
win.fn <- function (n,win.snps=100) {
    sites <- list( chr1=bcf.sites[[1]][seq(1+(n-1)*win.snps,length.out=win.snps)] )
    regions <- enclosing_region(sites)
    vcf_query( bcf.file, regions, get.indivs )
}
win.fn(20,10)  # get the 20th window of 10 SNPs
```

@@

Now let's see how it goes to do this for an entire chromosome.
On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
First, windows by SNPs.
```{r lpca_chr1_snp, cache=TRUE}
sites <- vcf_positions(bcf.file)
do.indivs <- sample( samples$ID, 50 )
win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=do.indivs) 
system.time( snp.pca <- eigen_windows(win.fn.snp,k=2) )
```
It takes about 2 minutes to do chromosome 1, for ten individuals, on phoebe.
Next, we compute the distance between windows with `pc_dist()` and do MDS visualization on the result.
This is just with ten samples, so we don't expect a real result, but let's check it all works:
```{r mds_chr1_snp, cache=TRUE, depends="lpca_chr1_snp"}
system.time( chr1.pcdist <- pc_dist( snp.pca ) )
# there may be windows with missing data
na.inds <- is.na( snp.pca$values[1,] )
chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
mds.coords <- chr1.mds$points[ ifelse( na.inds, NA, cumsum(!na.inds) ), ]
colnames(mds.coords) <- paste("MDS coordinate", 1:ncol(mds.coords))
```

@@

```{r plot_pc_corners}
countries <- samples$Country.of.Origin[match(do.indivs,samples$ID)]
country.names <- unique(countries)
country.cols <- rainbow_hcl( length(country.names), c=90 )
country.pch <- rep(1:6,length.out=length(country.names))
for (k in 1:ncol(corner.pcs$vectors)) {
    xy <- matrix( corner.pcs$vectors[,k], nrow=nrow(corner.pcs$vectors)/nrow(corner.pcs$values) )
    plot(xy, xlab="PC 1", ylab="PC 2", main=paste("corner",k), 
            col=country.cols[match(countries,country.names)], 
            pch=country.pch[match(countries,country.names)]  )
}
legend('topright', col=country.cols, pch=country.pch, legend=country.names)
```

>>>>

@@

To get windows out of the VCF file and recode them, we'll just pipe a call to `bcftools`.
Here's a dummy version that will pull out windows of 100 SNPs for only the first four individuals.
```{r win_fn}
get.indivs <- c("HM001","HM002","HM003","HM004")
bcf.file <- "data/chr1-filtered-set-2014Apr15.bcf"
# first read in the info about sites
bcf.con <- pipe(paste("bcftools query -f '%POS\\n' -r chr1:1-100000",bcf.file),open="r")
bcf.sites <- list( chr1=scan(bcf.con) )
close(bcf.con)
# now use bcftools to pull out sites we want
win.fn <- function (n,win.snps=100) {
    regions <- data.frame( chrom="chr1", start=1+(n-1)*win.snps, end=n*win.snps )
    vcf_query( bcf.file, regions, get.indivs )
}
win.fn(20,10)  # get the 20th window of 10 SNPs
```

@@

Now let's see how it goes to do this for an entire chromosome.
On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
First, windows by SNPs.
```{r lpca_chr1_snp, cache=TRUE}
sites <- vcf_positions(bcf.file)
do.indivs <- sample( samples$ID, 50 )
win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=do.indivs) 
system.time( snp.pca <- eigen_windows(win.fn.snp,k=2) )
```
It takes about 2 minutes to do chromosome 1, for ten individuals, on phoebe.
Next, we compute the distance between windows with `pc_dist()` and do MDS visualization on the result.
This is just with ten samples, so we don't expect a real result, but let's check it all works:
```{r mds_chr1_snp, cache=TRUE, depends="lpca_chr1_snp"}
system.time( chr1.pcdist <- pc_dist( snp.pca ) )
# there may be windows with missing data
na.inds <- is.na( snp.pca[1,] )
chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
mds.coords <- chr1.mds$points[ ifelse( na.inds, NA, cumsum(!na.inds) ), ]
colnames(mds.coords) <- paste("MDS coordinate", 1:ncol(mds.coords))
```

@@

```{r plot_pc_corners}
countries <- samples$Country.of.Origin[match(do.indivs,samples$ID)]
country.names <- unique(countries)
country.cols <- rainbow_hcl( length(country.names), c=90 )
country.pch <- rep(1:6,length.out=length(country.names))
for (k in 1:nrow(corner.pcs)) {
    xy <- matrix( corner.pcs[k,-(1:3)], ncol=2 )
    plot(xy, xlab="PC 1", ylab="PC 2", main=paste("corner",k), 
            col=country.cols[match(countries,country.names)], 
            pch=country.pch[match(countries,country.names)]  )
}
legend('topright', col=country.cols, pch=country.pch, legend=country.names)
```

### 2016-05-13 medicago/summarize_run.Rmd
added in total variance output to eigen_windows and split out cov_pca

Here are the extreme windows in the MDS plot:
```{r get_corners}
mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
# set up colors and pchs for corners
corner.cols <- brewer.pal(3,"Dark2")
corner.pch <- c(15,17,19)
ccols <- rep("black",nrow(mds.coords))
cpch <- rep(20,nrow(mds.coords))
for (k in 1:ncol(mds.corners)) {
    ccols[ mds.corners[,k] ] <- corner.cols[k]
    cpch[ mds.corners[,k] ] <- corner.pch[k]
}
# plot corners and MDS along the chromosome
layout(matrix(c(1,1,2,3),nrow=2),widths=c(1,2))
plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
     col=adjustcolor(ccols,0.75), 
     xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
for (k in mds.cols) {
    chrom.plot( mds.coords[,k], pch=20, 
            xlab="Position (Mb)", main=paste("MDS coordinate",match(k,mds.cols)),
            ylab=colnames(mds.coords)[k],
            col=adjustcolor(ccols,0.75) )
}
```

>>>>

Here are the extreme windows in the MDS plot:
```{r get_corners}
mds.corners <- corners( mds.coords[,mds.cols[1:2]], prop=.05 )
# set up colors and pchs for corners
corner.cols <- brewer.pal(3,"Dark2")
corner.pch <- c(15,17,19)
ccols <- rep("black",nrow(mds.coords))
cpch <- rep(20,nrow(mds.coords))
for (k in 1:ncol(mds.corners)) {
    ccols[ mds.corners[,k] ] <- corner.cols[k]
    cpch[ mds.corners[,k] ] <- corner.pch[k]
}
```
```{r plot_corners}
# plot corners and MDS along the chromosome
layout(matrix(c(1,1,2,3),nrow=2),widths=c(1,2))
plot( mds.coords[,mds.cols[1:2]], pch=cpch, 
     col=adjustcolor(ccols,0.75), 
     xlab="MDS coordinate 1", ylab="MDS coordinate 2" )
for (k in mds.cols) {
    chrom.plot( mds.coords[,k], pch=20, 
            xlab="Position (Mb)", main=paste("MDS coordinate",match(k,mds.cols)),
            ylab=colnames(mds.coords)[k],
            col=adjustcolor(ccols,0.75) )
}
```
